Estimating dark matter halo concentration using the integrated mass profile

### 2015-11-06 paper/paper.tex
intro revisada

In a dark matter dominated Universe galaxy properties are driven by
dark matter halo characteristics. 
One of the key structural halo properties is the concentration. 
[EN QUE MODELOS IMPORTA LA CONCENTRACION?]
For these reasons there has been a lot of effort to have a reliable
concentration estimate from N-body simulations.
Here we show that the concentration parameter in dark matter halos from
N-body simulations can be reliably estimated from the radially integrated mass
profile.
The main advantage of this approach is that it uses the full
particle information without any binning, which makes it reliable in
cases when low numerical resolution becomes a limitation for other
methods.   
We show the good performance of this method by estimating halo
concentration both on mock and N-body halos. 
We compare these results against two other methods:
maximum radial velocity measurements and radial particle binning to
fit the density profile. 
Tests on  the mock halos show that the accuracy of our method to recover input
concentrations varies with halo resolution. For halos
sampled with $20$ particles our method recovers the input
concentration with $10\%$ accuracy. 
For the maximum radial velocity method it is on the order of $20\%$
while density profile methods perform worst at the low resolution
limit since they require at least 200 particles inside the virial
radius to make a reliable estimate of the density profile. 
For halos with $2\times 10^4$ particles our method achieves an
accuracy of $0.01\%$ while the velocity and density profile methods
achieve $0.1\%$ and $1\%$, respectively. 
We measure the mass-concentration relationship on N-body
data taking care of using halos sampled with at least $200$
particles. We find that at low masses $10^{12}\hMsun <
M<10^{13}\hMsun$ our method yields median concentration values lower
by a $15\%-20\%$ compared to the velocity and density methods ({\bf
  why?}).  
At higher masses $3\times 10^{13}\hMsun < M < 2\times
10^{14}\hMsun$ the three methods give similar results.
\end{abstract}

@@

{\bf Navarro et.al. 2010 have used high resolution simulations to show
  that the full halo density profile is not universal and follows an
  Einasto density distribution. However, if one is not interested in
  the very central region of the dark matter halo (where galaxy
  formation takes place, and where the effects of baryon physics on
  dark matter distribution is still an uncknown), one can assume an
  approximation where dark matter overdensities closely follow an
  universal density profile. In a first approximation this profile is
  spherically symmetric and its density only dependens on the radial
  coordinate. Although it is well known that the characteristics of
  this densty profile depends on redshift (Maccio et.al. 2008), its
  universal shape seems to be independent of the cosmological
  parameters and is self-similar for different spatial scales after an
  adequate re-scaling is applied. \citep{NFW,Taylor2001}
}

{\bf In such circumstances}, one the most popular parameterizations
for dark matter halo radial density distribution is the
Navarro-Frenk-White (NFW) profile \citep{NFW}.  This profile is a
double power law in radius, where the transition break happens at the
so-called scale radius, $r_s$.  The ratio between the scale radius and
the virial radius $R_v$ is known as the concentration $c=R_v/r_s$.
Simulations show that the concentration is a strong function of halo
mass and redshift and provides information about the mass assembly of
the dark matter halo (Citar paper que reproduce mis graficas y conecta
C con m(z)).

Observationally, the relationship between halo mass and concentration
could provide a potential test of the Cold Dark Matter (CDM) paradigm
on galactic scales.  For this motivation a great deal of effort has
been invested in calibrating this relationship with simulations
\citep{Neto2007,Duffy2008,Munoz2011,Prada2012,Ludlow2014} and finding
the best possible way to constraint it with observations
\citep{Buote2007,Comerford2007,Mandelbaum2008,Giocoli2014,Foex2014,Shan2015}.

In N-body simulations there are two main methods to estimate the
concentration parameter of a dark matter halo in a N-body simulation.
The first method takes the particles composing the halos and bins them
in logarithmic radii to estimate the density in each bin, then it
proceeds to fit the density as a function of the radius to the NFW
profile.  A second method uses an analytic property of the NFW profile
that relates the maximum of the ratio of the circular velocity to the
virial velocity.  The concentration can be then found as the root of a
algebraic equation dependent on this maximum value.

@@

In this paper we propose a new method to estimate the dark matter halo
concentration in halos from N-body simulations.  Our method builds the
cumulative mass profile from the particle to find the best possible
concentration value fitting this distribution.  This proposal has two
advantages with respect to the two methods mentioned above.  It does
not involve any data binning and does not throw away data points {\bf
  resulting in a robust estimate even at low resolution/particle
  numbers. In the same sence, since the method does not require any
  binning, there is no need to tune numerical parameters, which
  provide definitive/independent estimates for the concentration
  parameter.}.

{\it mi opinión es que el método es muy bueno, y sobresalta sobre los
  otros dos en sus ventaja: robusto, no requiere bineo, aplica bien en
  resoluciones pobres. Como herramienta (que es lo que se está
  vendiendo en el trabajo, un método, eso es lo que hay que recalcar y
  resaltar)}

>>>>

In a dark matter dominated universe galaxy properties are driven by
the dark matter halo host.
One of the key structural halo properties is the concentration. 
[EN QUE MODELOS IMPORTA LA CONCENTRACION?]
For these reasons it is important to have a reliable concentration
estimate from N-body simulations. 
Here we show that the concentration parameter in dark matter halos from
N-body simulations can be reliably estimated from the radially integrated mass
profile.
The main advantage of this approach is that it uses the full
particle information without any binning, which makes it reliable in
cases when low numerical resolution becomes a limitation for other
methods.   
We show the good performance of this method by estimating halo
concentration both on mock and N-body halos. 
We compare these results against two other methods:
maximum radial velocity measurements and radial particle binning to
fit the density profile. 
Tests on  the mock halos show that the accuracy of our method to recover input
concentrations varies with halo resolution. For halos
sampled with $20$ particles our method recovers the input
concentration with $10\%$ accuracy. 
For the maximum radial velocity method it is on the order of $20\%$
while density profile methods perform worst at the low resolution
limit since they require at least 200 particles inside the virial
radius to make a reliable estimate of the density profile. 
For halos with $2\times 10^4$ particles our method achieves an
accuracy of $0.01\%$ while the velocity and density profile methods
achieve $0.1\%$ and $1\%$, respectively. 
We measure the mass-concentration relationship on N-body
data taking care of using halos sampled with at least $200$
particles. We find that at low masses $10^{12}\hMsun <
M<10^{13}\hMsun$ our method yields median concentration values lower
by a $15\%-20\%$ compared to the velocity and density methods ({\bf
  why?}).  
At higher masses $3\times 10^{13}\hMsun < M < 2\times
10^{14}\hMsun$ the three methods give similar results.
\end{abstract}

@@

{\bf FALTA CITEP Navarro et.al. 2010} have used high resolution simulations to show
that the full halo density profile is not universal and follows an
Einasto density distribution. 
However, if one is not interested in the very central region of the
dark matter halo (where galaxy formation takes place, and where the
effects of baryon physics on dark matter distribution is still an
uncknown), one can assume an approximation where dark matter
overdensities closely follow an universal density profile. 
In a first approximation this profile is spherically symmetric and
its density only dependens on the radial coordinate. 
Although it is well known that the characteristics of
this densty profile depend on redshift {\bf FALTA CITEP (Maccio
  et.al. 2008)}, its universal shape seems to be independent of the
cosmological parameters and is self-similar for different spatial
scales after an adequate re-scaling is applied \citep{NFW,Taylor2001}.

In such circumstances, one the most popular parameterizations
for dark matter halo radial density distribution is the
Navarro-Frenk-White (NFW) profile \citep{NFW}.  
This profile is a double power law in radius, where the transition
break happens at the so-called scale radius, $r_s$.  
The ratio between the scale radius and the virial radius $R_v$ is
known as the concentration $c=R_v/r_s$. 
Simulations show that the concentration is a strong function of halo
mass and redshift and provides information about the mass assembly of
the dark matter halo {\bf (Citar paper que reproduce mis graficas y conecta
C con m(z))}.

{\bf usos de modelos de formacion de galaxias}
The relationship between halo mass and concentration
could also provide a potential test of the Cold Dark Matter (CDM)
paradigm.

For all these reasons mentioned above a great deal of effort has
been invested in calibrating this relationship with simulations
\citep{Neto2007,Duffy2008,Munoz2011,Prada2012,Ludlow2014} and finding
the best possible way to constraint it with observations
\citep{Buote2007,Comerford2007,Mandelbaum2008,Giocoli2014,Foex2014,Shan2015}.

In N-body simulations there are two main methods to estimate the
concentration parameter of a dark matter halo.
The first method takes the halo particles and bins them into
logarithmic radii to estimate the density in each bin, then it 
proceeds to fit the density as a function of the radius to the NFW
profile.  
A second method uses an analytic property of the NFW profile
that relates the maximum of the ratio of the circular velocity to the
virial velocity.  The concentration can be then found as the root of a
algebraic equation dependent on this maximum value.

@@

In this paper we show that integrated mass profile can be used  to
estimate the dark matter halo concentration in halos from N-body
simulations.  
This has two advantages with respect to the two methods mentioned
above.  
It does not involve any data binning and does not throw away data
points resulting in a robust estimate even at low resolution/particle
numbers.  
Furthermore, since the method does not require any binning,
there is no need to tune numerical parameters, providing a new
independent method to estimate the concentration parameter.
